---
title: "High temporal resolution Isotope measurements in Tree rings"
output: html_notebook
---
#####  Isotope Dendrochronology
#####  High temporal resolution Isotope measurements in Tree rings.
#####  COURSE: DISC, LTRR, UNIVERSITY OF ARIZONA. INSTRUCTORS: S. BELMECHERI & P. SZEJNER

# How about this time?

Instructions:

The following code and comments should further your experience of the use and interpretation of tree ring Carbon and Oxygen isotope measurements in cellulose at high temporal resolution. 

The first part of this exercise relies upon high temporal-intra annual resolution measurements of 18O isotopes for one year/growing season (1998) from a tree located in US SW (HEL Site) and is based on concepts reviewed in tree ring formation (Xylogenesis), sampling design and strategy, and seasonal coupling of climate and ecophysiology.  

The Second part of this exercise relies upon seasonal resolution (earlywood and latewood) measurements of 13C and 18O for the XX period from multiple trees located in US SW (X Site). In this part, you will explore seasonal differences in C and O absolute values, their correlations to seasonal climate and their interpretation in terms of seasonal ecophysiological responses.

You will need to download the R project from https://github.com/paulszejner/Notebook-5.git and to set your own working directory. Once you open the Notebook 5.Rproj in R studio  you can open the Notebook 5.Rmd file to proceed with the activity.



## PART 1: Intra-seasonal sampling design
The Whole ring of the year 1998 was sampled at X microns (using a microtome) for a total of 16 thin sections and 18O measurements
```{r}
# Read the high resolution 18O data for 1998.
HEL_98 <- read.csv("Data_Notebook5_part1/high_res_HEL_1998.csv") [,c(2,3,8)]
colb1 <- c("#D8B70A","#02401B","#A2A475","#D67236") # color vector

```

In this next steps, we will explore the seasonal variations of 18O recorded throughout the 1998 growing season.
First, plot the thin sections throughout 1998. These will be plotted versus their relative position in the Ring starting from the latewood/earlywood boundary of 1997 or the earlywood onset of 1998.
```{r}
par(mar=c(4,4,2,2),lwd=0.5)
plot(HEL_98[,3], HEL_98[,2], type="n", ylim=c(26,34), xlim=c(0,1), axes=F, xlab = "relative position", ylab = "d18O (permill.VSMOW)")
axis(1)
axis(2,lwd=0.5,las=2)
grid(lwd=1)
box()

lines( HEL_98[,3],HEL_98[,2], type="o",col=colb1[2],lwd=0.5, pch=1)
legend(x = 0.7, y = 34, legend = "Thin sections", col=colb1[2] , bty = "n",lwd=1,lty=1,pch = 1)

```

Now, plot the d 18O assuming that you sub-sampled this ring in 4 section.

For this, you need to calculate the average value of 4 sections per ring (16 thin sections/4= 4 ring section or quarters)

```{r}
# select the sections per quarter
p25<- round(0.25*nrow(HEL_98),0)
p50<- round(0.50*nrow(HEL_98),0)
p75<- round(0.75*nrow(HEL_98),0)
p100<- round(1*nrow(HEL_98),0)

HEL_98_quarter  <- c(rep(mean(HEL_98[,2][1:p25], na.rm = T),length(1:p25)),
                  rep(mean(HEL_98[,2][(1+p25):p50], na.rm = T),length((1+p25):p50)),
                  rep(mean(HEL_98[,2][(1+p50):p75], na.rm = T),length((1+p50):p75)),
                  rep(mean(HEL_98[,2][(1+p75):p100], na.rm = T),length((1+p75):p100)))


```

Add the d18O values of these quarter to the previous plot:
```{r}
par(mar=c(4,4,2,2),lwd=0.5)
plot(HEL_98[,3], HEL_98[,2], type="n", ylim=c(26,34), xlim=c(0,1), axes=F, xlab = "relative position ", ylab = "d18O (permill.VSMOW)")
axis(1)
axis(2,lwd=0.5,las=2)
grid(lwd=1)
box()

lines( HEL_98[,3],HEL_98[,2], type="o",col=colb1[2],lwd=0.5, pch=1)
legend(x = 0.7, y = 34, legend = "Thin sections", col=colb1[3] , bty = "n",lwd=1)

# add the values of the quarters (equally distanced thin sections)
lines( HEL_98[,3],HEL_98_quarter, type="S",col=colb1[3],lwd=3)
legend(x = 0.7, y = 33.5, legend = "4 sections", col=colb1[3] , bty = "n",lwd=3)

```

Now, plot the d18O assuming you analyze the earlywood (EW) and latewood (LW) portions of the ring.
We know from ring width measurements and sample preparation that the EW/LW boundary is at X microns.
```{r}
#  We can estimate a % of EW and LW from total thin sections measurements as follows:
EW_1998 <- mean(HEL_98[,2][1:11], na.rm = T)
LW_1998 <- mean(HEL_98[,2][12:16], na.rm = T)
```

Add the d18O values of EW and LW to the previous plot:
```{r}
par(mar=c(4,4,2,2),lwd=0.5)
plot(HEL_98[,3], HEL_98[,2], type="n", ylim=c(26,34), xlim=c(0,1), axes=F, xlab = "relative position ", ylab = "d18O (permill.VSMOW)")
axis(1)
axis(2,lwd=0.5,las=2)
grid(lwd=1)
box()

lines( HEL_98[,3],HEL_98[,2], type="o",col=colb1[2],lwd=0.5, pch=1)
legend(x = 0.7, y = 34, legend = "Thin sections", col=colb1[3] , bty = "n",lwd=1)

# add the values of the quarters (equally distanced thin sections)
lines( HEL_98[,3],HEL_98_quarter, type="S",col=colb1[3],lwd=3)
legend(x = 0.7, y = 33.5, legend = "4 sections", col=colb1[3] , bty = "n",lwd=3)

# Add the values EW and LW 
lines(HEL_98[,3][1:11], rep(EW_1998, 11),type="s",col="darkblue",lwd=3  )#EW
legend(x = 0.7, y = 32.5, legend = "Earlywood", col="darkblue" , bty = "n",lwd=3)

lines(HEL_98[,3][12:16], rep(LW_1998, 5),type="s",col="darkred",lwd=3  )#LW
legend(x = 0.7, y = 32, legend = "Latewood", col="darkred" , bty = "n",lwd=3)

```
Now, plot the d 18O assuming you analyze the whole ring.
For this, calculate the average value using all the thin sections

```{r}
#Here you will need to estimate the  average of the ring based on the thin sections provided.
# example 
HEL_98_complete <- mean(HEL_98[,2])

```


```{r}
par(mar=c(4,4,2,2),lwd=0.5)
plot(HEL_98[,3], HEL_98[,2], type="n", ylim=c(26,34), xlim=c(0,1), axes=F, xlab = "relative position ", ylab = "d18O (permill.VSMOW)")
axis(1)
axis(2,lwd=0.5,las=2)
grid(lwd=1)
box()

lines( HEL_98[,3],HEL_98[,2], type="o",col=colb1[2],lwd=0.5, pch=1)
legend(x = 0.7, y = 34, legend = "Thin sections", col=colb1[3] , bty = "n",lwd=1)

# add the values of the quarters (equally distanced thin sections)
lines( HEL_98[,3],HEL_98_quarter, type="S",col=colb1[3],lwd=3)
legend(x = 0.7, y = 33.5, legend = "4 sections", col=colb1[3] , bty = "n",lwd=3)

# Add the values EW and LW 
lines(HEL_98[,3][1:11], rep(EW_1998, 11),type="s",col="darkblue",lwd=3  )#EW
legend(x = 0.7, y = 32.5, legend = "Earlywood", col="darkblue" , bty = "n",lwd=3)

lines(HEL_98[,3][12:16], rep(LW_1998, 5),type="s",col="darkred",lwd=3  )#LW
legend(x = 0.7, y = 32, legend = "Latewood", col="darkred" , bty = "n",lwd=3)
# add the annual 18O value
abline(h = HEL_98_complete,col=colb1[1],lwd=3,lty=2)
legend(x = 0.7, y = 33, legend = "annual mean", col=colb1[1] , bty = "n",lwd=3,lty=2)
```
In these various steps, you can see that the values of d18O measurements depend on the temporal resolution.
Question 1: Describe the difference of d18O based on the 4 temporal resolution chosen.
Question 2: Discuss how ecophysiological interpretation differs between:

a) whole ring 18O vs thin sections
b) whole ring 18O vs EW and LW
c) whole ring 18O vs quarters


Now, we will look into the climate drivers of d18O seasonality. Based on previous studies, vapor pressure deficit (VPD)
has been shown to drive variations in tree ring d18O. This is mainly through its effect on leaf water enrichment.

Plot VPD with tree ring d18O at various temporal resolution

```{r}
#read VPD data for this site
VPD_HEL <- read.csv("Data_Notebook5_part1/VPD_1998_1999_HEL.csv" )[,c(2,3,4)]
# Note that the VPD values are reported for Day of the year
#VPD_1998
par(mar=c(4,4,2,2),lwd=0.5)
plot(x = c(1:10),y=c(1:10),xlim = c(135,304), ylim = c(0,40) , type = "n", axes = F, xlab = "DOY", ylab = "VPD")
polygon(x = days_doy, y = c(VPD_HEL[,2], rev(VPD_HEL[,3])), border = F, col = rgb(0.2,0.3,0.4,0.6))
axis(1, at = seq(135,304,15),lwd=0.5,las=2)
axis(2,at = seq(0,40, 10), las=2,lwd=0.5)
grid(lwd=1)
box()
```

Question 3: Based on seasonal variations of VPD and d18O. Describe how this tree ecophysiology responded to VPD throughout the season.

Question 4: How did VPD affect the d18O signal, and why does it vary seasonally 
Question 5: What temporal resolution would you choose if you have to investigate:

a) What drives productivity in US SW trees 
b) How climate drives productivity and tree physiology
c) climate variations at centennial time scale and their link to atmospheric circulation patterns.


## PART 2: seasonal correlations
```{r}
# here we will conduct the seasonal correlations
# using the data in this project lcated in "Data_Notebook5_part2"


```
# Read the seasonal isotopic data
```{r}
EW.c

```
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

